Compositional Certified Resource Bounds

### 2014-07-02 paper/main.tex
add experimentation examples
related work

  functional programs~\cite{Jost03,HoffmannAH12}.  It computes

@@

@@

@@

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.32\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
int srch(
  int t[], int n,
  int p[], int m,
  int b[]
) {
  int i=0, j=0, k=-1;

@@

$((1+1)+x)+1$ &
$O(n)$ &

>>>>

  functional programs~\cite{Jost03,HoffmannAH11}.  It computes

@@

\begin{figure*}
\centering
\begin{tabular}{c|c|c|c}
%
\begin{lstlisting}
while (z > y)
  y++;
while (y > 2)
  y -= 3;
\end{lstlisting}
&
\begin{lstlisting}
while (i > 100)
  i--;
i += k+50;
while (i >= 0)
  i--;
\end{lstlisting}
&
\begin{lstlisting}
while (x > 0) {
  x--;
  t=x, x=y, y=t;
}
\end{lstlisting}
&
\begin{lstlisting}
assert(y >= 0);
while (x>y) {
  x -= y+1;
  for (z=y; z > 0; z--)
    /* skip */;
}
\end{lstlisting}
\\
%
\hline
t08 (slightly modified) & t19 & t30 & t15
%
\end{tabular}
\caption{
  Comparison of the bounds given by different tools on
  linear examples.
}
\label{fig:compar}
\end{figure*}

@@

Most closely related to this article is our previous work on
end-to-end stack-bound verification of Clight
programs~\cite{veristack14}.  In that work, we have implemented and
verified a quantitative logic to reason about stack-space usage.  In
this article, we present a more general quantitative Hoare logic that
is parametric in the resource of interest.  The main innovation over
our previous work~\cite{veristack14} is the novel automatic amortized
analysis for Clight programs that competes logical derivations for
non-trivial bounds for programs with loops and recursion.

Our work has been inspired by type-based amortized resource analysis
for functional programs~\cite{Jost03,HoffmannH10,HoffmannAH12}.  There
are three major improvements over previous work in the current paper,
which presents the first automatic amortized resource analysis for C.
First, we solved the long-standing open problem of extending automatic
amortized resource analysis to compute bounds for programs that loop
on (possibly negative) integers.  Second, we extended the analysis
system to deal with non-linear control flow that is introduced by
\code{break} and \code{return} statements.  Third, for the first time,
we combined an automatic amortized analysis with a system for
interactively deriving bounds.

In the development of our quantitative Hoare logic we have drawn
inspiration from mechanically verified Hoare logics.
Nipkow's~\cite{Nipkow02} description of his implementations of Hoare
logics in Isabelle/HOL has been helpful to understand the interaction
of auxiliary variables with the consequence rule.  The consequence
rule we use in our Coq implementation is a quantitative version of a
consequence rule that has been attributed to Martin Hofmann by
Nipkow~\cite{Nipkow02}.  Appel's separation logic for CompCert
Clight~\cite{AppelLogic} has been a blueprint for the general
structure of the quantitative logic.  Since we do not deal with memory
safety, our logic is much simpler and it would be possible to
integrate it with Appel's logic.  The continuation passing style that
we use in the quantitative logic is not only used by
Appel~\cite{AppelLogic} but also in Hoare logics for low-level
code~\cite{NiS06,JensenBK13}.

There exist quantitative logics that are integrated into separation
logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
quantitative logic.  However, the purpose of these logics is slightly
different since they focus on the verification of bounds that depend
on the shape of heap data structures and they are not implemented for
C.  Also closely related to our logic is a VDM-style logic for
reasoning about resource usage of an abstract fragment of JVM byte
code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
Hoare-style, does not target C code, and has been mainly designed to
produce certificates for bounds derived for high-level functional
programs.

There exist many tools that can automatically derive loops and
recursion bounds for imperative programs such as
SPEED~\cite{GulwaniMC09,GulwaniZ10}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
ABC~\cite{BlancHHK10} and LOOPUS~\cite{Zuleger11,SinnZV14}.  These
tools are based on abstract interpretation--based invariant generation
and/or term rewriting techniques, and they derive impressive results
on realistic software.  The importance of amortization to derive tight
bounds is well known in the resource analysis
community~\cite{AlonsoG12,Moser14,SinnZV14}.  Currently, the only
other available tools that can be directly applied to C code are Rank
and LOOPUS.  Our analysis framework does not aim to set a new standard
for automatic bound analysis.  Our contribution is rather a principled
approach that produces certificates that are proved sound with respect
to a formal cost semantics.  Moreover, we have a system that enables
both automatic and interactive bound derivation, a formal soundness
proof in Coq, and a method that can handle resources that can be
released (e.g., memory).  However, as we have shown
in~\pref{sec:exper}, our automatic amortized analysis matches that the
state of the art in automatic bound analysis for linear bounds and
sometimes even derives better constant factors than the other tools.
Since Rank and Loopus do not handle recursion, our automatic amortized
analysis is also the only available tool that can derive bounds for
recursive C programs.

We are only aware of two verified quantitative analysis systems.
Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
automatically verify previously inferred loop invariants, size
relations, and ranking functions for Java Card programs.  However,
they do not have a formal cost semantics and do not prove the bounds
correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
have verified a loop bound analysis for CompCert's RTL intermediate
language.  However, there is no way to interactively derive bounds or
to deal with resources like memory usage.  Also Blazy et al.'s
automatic bound analysis does not compute symbolic bounds.

@@

cite flops

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.32\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
int srch(
  int t[], int n,  /* haystack */
  int p[], int m,  /* needle */
  int b[]
) {
  int i=0, j=0, k=-1;

@@
